---
title: 'Level 1 Data Cleaning: Clean the mtcars Dataset'
---

# Objective

The objective of this assignment is to practice cleaning and transforming a messy dataset using tidyverse functions. You will use skills like renaming and reordering columns, sorting rows, changing data types, mutating data, and using the stringr and forcats packages.

In this Level 1 Cleaning assignment, you will work with a simple dataset and focus on basic data cleaning tasks. Most tasks are outlined in the assignment script. You may want to review the [Data Cleaning Walkthrough]() before you begin.

You may additionally or alternatively complete the [Level 2 Data Cleaning assignment](). In Level 2, you will work with a more complex dataset and perform additional cleaning tasks with less direct instruction. The Level 2 assignment has more opportunities to demonstrating meeting course standards than this Level 1 assignment and is recommended for those who are already comfortable with the tasks in this assignment.

# Instructions

1. If you have not already done so, pull the latest changes from the `d2mr-assignments` repository to ensure you have the most up-to-date version of the assignment files. Confirm you are working in your fork of the repository.
2. Open `cleaning-level-1.qmd` in RStudio and follow the instructions in the Setup section below to load and inspect the (clean) `goal.mtcars` dataset. 
    - **Important:** The provided code makes a small modification to the original `mtcars` dataset to create a `goal.mtcars` dataset. You will use this goal dataset as a reference to clean the `messy-mtcars.csv` dataset, not the true original.
3. Follow the tasks described in the assignment script to clean the dataset and return it to its original state.
4. At several points in this document you will come across questions or non-coding exercises. Answer these questions in the text of this .qmd document, immediately below the question.
5. *Optional:* Continue to follow the instructions in the assignment script to clean the dataset above and beyond matching the original. 


### Tasks

**Reminder:** Your goal is to match `goal.mtcars` (created below), *not* the default `mtcars`.[^1]

[^1]: Why? The `mtcars` dataset includes models names as *row names* rather than as a variable/column. For the purposes of this cleaning exercise, it's more useful to treat model names as a variable.

1. **Renaming Columns:** Rename columns to match the original `mtcars` dataset column names.
2. **Reordering Columns:** Ensure the columns are in the correct order.
3. **Sorting Rows:** Sort the rows by `mpg` (miles per gallon) and then by `cyl` (number of cylinders).
4. **Changing Data Types:** Ensure `gear` and `carb` are factors and other columns are numeric.
5. **Mutating Data:** Create a new column `hp_per_cyl` calculated as `hp` (horsepower) divided by `cyl`.
6. **String Manipulation:** Use `stringr` to clean any unwanted spaces or characters in the `model` column.
7. **Handling Factors:** Use `forcats` to ensure the levels of `gear` are ordered as `3`, `4`, `5`.

# Setup

## Loading libraries and set seed

```{r}
#| label: setup
library(tidyverse)
set.seed(1234)
```


## Read in and inspect messy data

Read in and inspect the messy dataset `messy-mtcars.csv`.

```{r}

#| label: read-messy-data

### LEAVE THIS CHUNK AS-IS ###

# You *might* need to edit the filepath, but don't change anything else!

# Read in messy-mtcars.csv
messy.mtcars <- read_csv(
  ########################################
  "messy-mtcars.csv", ## <-- THIS IS THE ONLY THING IN THIS CHUNK YOU CAN CHANGE IF NECESSARY
  ########################################
  trim_ws = FALSE, name_repair = "minimal", col_types = cols(.default = col_character()))

# Inspect the data
head(messy.mtcars)

```

## Inspect the original mtcars dataset

```{r}
#| label: inspect-original-data

### LEAVE THIS CHUNK AS-IS ###

# Load the original mtcars dataset
data(mtcars)

# Create the "goal.mtcars" dataset
# Convert row names to a column called "model" (see note above about row names)
goal.mtcars <- mtcars %>%
  rownames_to_column(var = "model")

# Inspect the goal.mtcars dataset
head(goal.mtcars)

# Optionally inspect the original mtcars dataset to see what the row names vs column issue looks like
# head(mtcars)

```


QUESTIONS:

1. What are the differences between the messy dataset and the original mtcars dataset?

- The original mtcars dataset is tidy but the messy datset is not.
- Columns are in a different order: In the messy dataset, the order of columns has been shuffled around. For example, "Car Model" is listed before "mpg," whereas in the original mtcars dataset, "mpg" comes first.
- Column names do not match: The names of the columns are changed in the messy dataset. For instance, "mpg" is labeled as "mpg (Miles/Gallon)" in the messy version, and "hp" is renamed to "horsepower."
-Data types are incorrect: The messy dataset treats everything as text, but the original mtcars dataset uses numeric values for columns like "mpg," "hp," and "wt," and categorical values for things like "cylinders" and "gear."
-Problems in the 'Car Model' column: The car names in the messy dataset have extra spaces, inconsistent capitalization, and potential typos, which isn’t an issue in the original dataset where everything is properly formatted.
-Commas instead of periods: Some numbers in the messy dataset (like in the "hp" and "wt" columns) use commas instead of periods, which should be fixed to ensure the data is correctly formatted.
-Inconsistent rounding: The messy dataset doesn’t round numbers like "drat" and "qsec" consistently, whereas the original dataset makes sure all the numbers are rounded to a similar number of decimal places.

2. What are the main issues you need to address in cleaning?
- Rename columns: Adjust column names to match the original dataset (e.g., change "mpg (Miles/Gallon)" back to "mpg" and "horsepower" back to "hp").
- Reorder columns: Arrange the columns in the same order as the original mtcars dataset for consistency.
- Change data types: Convert columns like "mpg," "hp," and "wt" to numeric, and columns like "cylinders" and "gear" to factors.
- Correct values in 'Car Model' column: Remove extra spaces, fix typos, and ensure consistent capitalization in the car model names.
- Remove commas from numbers: Replace commas with periods in numeric columns like "hp" and "wt" to make the data properly formatted for analysis.
- Round numbers to the same number of decimal places (1): Ensure that numbers in columns like "drat" and "qsec" are rounded to one decimal place for consistency.

# Clean the Dataset

## Creat "clean" dataset

```{r}

#| label: make-cleaning-dataset

# Create a dataset to work with during the cleaning process called "clean.mtcars"
clean.mtcars <- messy.mtcars

```


## Clean columns/variables

# Check the names and order of columns in iris:
colnames(mtcars)

# Check the names and order of columns in messy.iris:
colnames(messy.mtcars)

```{r}

#| label: rename-columns

# Rename columns to match the original mtcars dataset
clean.mtcars <- messy.mtcars %>%
  rename(mpg = `mpg (Miles/Gallon)`, cyl = cylinders, disp = dsip, hp = horsepower, drat = drat, wt = wt, qsec = qsec, vs = vs, am = am, gear = gearbox, carb = carburetors)

```

```{r}

#| label: reorder-columns

# Reorder columns to match the original mtcars dataset
clean.mtcars <- clean.mtcars %>%
  relocate(mpg, cyl, disp, hp, drat, wt, qsec, vs, am, gear, carb)

```

```{r}

#| label: correct-data-types

#222222 Correct variable data types

```

## Checkpoint 1

```{r}

#| label: checkpoint-1

# Inspect the current state of the dataset
head(clean.mtcars)

# Use all.equal() to check if clean.mtcars matches goal.mtcars
all.equal(clean.mtcars, mtcars)

```

Questions:

3. The current state of the dataset does not yet match the `goal.mtcars` dataset. Explain the issues that `all.equal()` flags.  Are there groups of issues that can be addressed together?

- The current state of the dataset does not yet match the goal.mtcars dataset, as all.equal() flags multiple issues. The main issue is that the columns in clean.mtcars are currently of the wrong data type; they are all being treated as characters, but they should be numeric or factors. Specifically, columns such as mpg, cyl, disp, hp, drat, wt, qsec, vs, am, gear, and carb should be numeric or factors, not characters. Additionally, there is a length mismatch, which suggests that the column order or the number of columns may differ between clean.mtcars and goal.mtcars. These issues can be addressed together by converting the columns to their correct data types and ensuring the columns are correctly ordered and named to match the goal.mtcars dataset.

3. Install (if necessary) and load the `daff` package. In the code chunk below, use the `diff_data()` function to identify the differences between `clean.mtcars` and `goal.mtcars`. 

```{r}
#| label: diff-data

# Install and/or load the daff package if needed

install.packages("daff")
library(daff)

# Use render_diff() and diff_data() to identify differences between clean.mtcars and goal.mtcars
render_diff(diff_data(clean.mtcars, mtcars))

```

2222222How is this method of identifying data discrepancies different from `all.equal()`? Which do you find more helpful at this point?

- The main difference is that all.equal() gives a simple yes/no answer or a short description of the differences, while diff_data() visually highlights the differences between datasets, making it much easier to see exactly where they differ.

- diff_data() is more helpful for visually understanding discrepancies, especially in larger datasets. all.equal() is quick and useful for a simple check but doesn’t provide detailed insights.

## Clean data values

Identifying specific value problems and cleaning them might take a little creativity at times! Use the troubleshooting/debugging workflows we've discussed to get as far as you can.

```{r}

#| label: remove-unwanted-spaces

# Remove unwanted whitespace in any columns required


```

```{r}

#| label: impute-missing-values

# Impute missing values for cyl and hp

```

```{r}

#| label: round-numeric-values

# Round numeric values

```

```{r}

#| label: fix-typos

# Fix typos in model names
## Hint: This can be completed with a 6-line mutate() statement



```

```{r}

#| label: remove-extra-column

# Remove the extra column


```


## Checkpoint 2

```{r}
#| label: checkpoint-2

# Inspect the dataset and compare to goal.mtcars

# all.equal()

# daff

```

Questions:

5. Is your dataset identical to `goal.mtcars`? If not, what are the remaining issues? If there were any issues you could not resolve in code, describe a proposed solution in plain English.

<!-- answer below -->


# Optional Cleaning Tasks

You can call it a day here, or continue with the optional cleaning tasks below. These tasks are a chance to challenge yourself and practice your skills further. They are also additional opportunities to demonstrate mastery of course standards! 


## Additional guided cleaning

*Optional:* Complete any number of the following tasks to further clean the dataset. Each task should be in its own code chunk with an appropriate label and clearly commented code:

1. Create a new factor column `mpg_category` that categorizes `mpg` into "low", "medium", and "high" based on the distribution of `mpg` values.
2. Create a new factor column `wt_category` that categorizes `wt` into "light", "medium", and "heavy" based on the distribution of `wt` values, then reverse the order of the levels.
3. Create a new column `mpg_per_cyl` that calculates `mpg` divided by `cyl`.
4. Filter the data to only include cars with automatic transmissions. (Hint: `?mtcars` will show you useful info.)
5. Identify which variable would be better suited as a logical variable, then convert it to logical.
6. Sort the dataset by `mpg_category` then reverse alphabetically by `model` (so that models with the same `mpg_category` are sorted Z to A).
7. Write the cleaned dataset as a csv file called `clean-mtcars.csv`, then read the csv back in maintaining the correct data types.


## Unguided cleaning and transformation

*Optional:* If you have the time and interest, continue transforming this dataset as you please. Create new columns based on the existing ones, reformat strings, try your hand at a regex replacement, summarize by groups (factor levels), visualize a simple relationship, or anything else you can think of. You can do this in addition to or instead of the suggested additional cleaning tasks above.


# Submission & Assessment

To submit:

1. Modify the `assessment.md` in this mini-project's directory:
    1. Check off all objectives you believe you have demonstrated
    2. Indicate which objectives you are meeting for the first time (if any)
    3. Complete any relevant open-ended items
2. Push your changes to your centralized assignment repository on GitHub. 
3. Confirm that Dr. Dowling nad your section TA are added as collaborators to your repository.
4. Submit your work in your next open mini-project assignment by including the following information in the text box:
    1. The title of the assignment: "Assignment: Clean the mtcars Dataset (Level 1)"
    2. A link to the **directory** for this assignment in your centralized assignment repo


